\section{Non-Uniform Binary Fuse Filters}

This project was completed for CSC 592: Algorithms for Big Data.

\subsection{Motivation}

% DEFINE U, N
Distributed systems often require fast and small set data structures. For example, content delivery networks avoid caching one-hit wonders by querying a visited
set~\cite{maggs2015}. Supporting such exact membership queries on a large universe\footnote{Set of all keys.} $U$ requires $\Omega(\lg |U|)$ bits per key. For 
large enough sets, this is impractical. By permitting false positive results with some probability $\epsilon$, the lower bound is reduced to $\lg \frac{1}{\epsilon}$ 
bits per key~\cite{carter1978}. In other words, \textit{approximate} membership queries can be much more memory efficient than exact membership queries. \\

\noindent
A binary fuse filter~\cite{graf2022} is a recent static probabilistic data structure for approximate membership queries. Membership information is stored in an 
array of $m$-bit values partitioned into disjoint segments. To query the filter, a key is hashed to a fingerprint value and to three array entries located in 
consecutive segments (see Figure~\ref{fig:csc592_bigdata_bff}). If the bitwise-XOR of the entries is equal to the fingerprint, then the query \emph{might} be in 
the set. Otherwise, the query \emph{definitely} is not in the set.

\input{figures/csc592-bigdata-bff.tex}

\noindent
To construct a binary fuse filter, choose each array entry such that, for each key, the bitwise-XOR of the three corresponding array entries is equal to
the fingerprint. This requires solving a system of linear equations in $\mathbb{F}^{m}_2$. Let $h_1$, $h_2$ and $h_3$ be the hash functions, $f$ be the 
fingerprint function, and $A$ be the array. Each key $k$ yields a linear equation $A[h_1(k)] \oplus A[h_2(k)] \oplus A[h_3(k)] = f(k)$ that must be satisfied 
(see Figure~\ref{fig:csc592_bigdata_soe}). If some equation is not satisfied, the filter will incorrectly report the corresponding key is \emph{definitely} not 
in the set. Since false negatives are not permitted, the choice of array entries must satisfy every key's equation.

\input{figures/csc592-bigdata-soe.tex}

\noindent
To solve the system, model it as a hypergraph. A hypergraph is a graph where edges may contain any subset of vertices. For each entry, create a vertex. For each 
equation, create an edge containing entries (vertices) in the equation (see Figure~\ref{fig:csc592_bigdata_hg}). Then, 2-peel the hypergraph, that is:
\begin{enumerate}
    \item Remove vertices of degree zero, then set their entries to $0^m$.
    \item Remove vertices of degree one and their incident edges. Then, set their entries to the bitwise-XOR of the fingerprint and the two other entries in the
    edge. If the other entries are not yet set, defer until they are set.
    \item Repeat until there are no vertices of degree less than two, or until the empty hypergraph is reached (i.e. all entries set).
\end{enumerate}
If the empty hypergraph is reached, the hypergraph is called 2-peelable. The system is solved if and only if its hypergraph is 2-peelable. \\

\input{figures/csc592-bigdata-hg.tex}

\noindent
In the filter setting, edges represent keys and vertices represent array entries. Higher edge density means lower space usage, since there are more keys (edges)
per array entry (vertex). Many families of random hypergraphs exhibit sharp phase transitions in peelability. When the edge density is below a critical value, 
called the 2-peelability threshold, sampled hypergraphs are almost surely 2-peelable. When the density exceeds this threshold, they are almost surely not. The 
threshold is therefore a direct measure of space efficiency. A higher threshold means more keys can be stored per array entry. \\

\noindent
Binary fuse filters induce the fuse graph family, which achieves the maximum 2-peelability threshold among 3-uniform hypergraph\footnote{In general, a 
hypergraph is $m$-uniform if every edge contains exactly $m$ vertices.} families. This threshold occurs at an edge density of $0.918$, or roughly 
$1.089 \lg \frac{1}{\epsilon}$ bits per key. No family of 3-uniform hypergraphs has higher 2-peelability threshold. More generally, a binary fuse filter with 
$m \geq 3$ hash functions\footnote{Only binary fuse filters with three or four hash functions are practical. With four hash functions, the 2-peelability 
threshold is already $0.977$, yielding about $1.023 \lg \frac{1}{\epsilon}$ bits per key or within $2.3\%$ of optimal.} saturates the 2-peelability threshold 
for $m$-uniform hypergraphs. \\

\noindent
However, this optimality result is confined to uniform hypergraphs. There may exist families of non-uniform hypergraphs with higher 2-peelability thresholds. If
such families also have a low average edge size (i.e. few hashes and accesses per query), filters based on them could potentially reduce memory overhead without
degrading query latency and throughput. Therefore, we ask:

% TODO: Add bold RQx
\begin{enumerate}
    \item How can we systematically construct non-uniform hypergraph families with peelability thresholds comparable to or surpassing those of fuse graph 
    families?
    \item How can we design efficient approximate membership query data structures by exploiting peelable non-uniform hypergraph families?
\end{enumerate}

\subsection{Summary}

\subsubsection{RQ1}

We computed 2-peelability thresholds for 10K non-uniform fuse graph families with mean edge size between 3 and 4 (see 
Figure~\ref{fig:csc592_bigdata_nuhg_graph}). As peelability is monotonic in edge density, we used binary search. At each candidate threshold, we sampled 
three hypergraphs with 1M edges. If the majority were peelable, we considered the family peelable at that threshold. We implemented this algorithm in C++ and 
computed thresholds in parallel on a multicore cluster. \\

\input{figures/csc592-bigdata-nuhg-graph.tex}

\noindent
While $m$-uniform graphs have integer mean edge sizes, they can be extended to any mean edge size via sharding. For example, splitting edges evenly between 
edge-disjoint 3-uniform and 4-uniform subgraphs yields a mean edge size of 3.5. Many non-uniform families outperform this baseline, and therefore afford 
smaller filters. Interestingly, some non-uniform families also outperform uniform families at integer mean edge sizes. Since peelability thresholds are defined 
in the limit of large graphs, a non-uniform family can be more peelable than a uniform family with nearly the same mean edge size. For example, one family 
achieves $\approx 1$\% memory savings over the uniform baseline at mean edge size $4$. \\

\noindent
These data suggest two conjectures. Let $\mathcal{H}_d$ denote the set of fuse graph families with mean edge size $d$ and let $\tau(\mathcal{F})$ 
denote the 2-peelability threshold of a family $\mathcal{F}$. Define $\Gamma(d) = \sup \{ \tau(\mathcal{F}) : \mathcal{F} \in \mathcal{H}_d \}$. We conjecture 
that $\Gamma(d)$ is continuous, monotonically increasing, and concave. Since a concave $\Gamma$ would lie above any linear interpolation, this implies that 
sharding is suboptimal. We further conjecture that for $d \not \in \mathbb{Z}$, the supremum is attained by the family with a fraction $(1 - p)$ of 
$\lfloor d \rfloor$-edges and a fraction $p$ of $\lceil d \rceil$-edges, where $p = d - \lfloor d \rfloor$. For example, for $d = 3.5$, this family contains
50\% 3-edges and 50\% 4-edges. Both conjectures remain open.

\subsubsection{RQ2}

\noindent
We implemented and optimized a binary fuse filter based on non-uniform hypergraph with 50\% 3-edges and 50\% 4-edges in C++, since this combination afforded the
greatest gains over sharding uniform binary fuse filters. Each key is hashed to either three or four array locations. Naively,
one might extract a bit from the hash and conditionally XOR a fourth array entry. However, since this bit comes from the hash, the branch is unpredicable and
is taken with 50\% probability. The branch can be eliminated by always fetching the fourth array entry and then zeroing it with a multiply. However, this places
additional pressure on the memory system. Instead of averaging 3.5 cache misses for large filters, this scheme averages 4. While this does not strongly impact 
latency due to memory level parallelism, it harms throughput. Instead, the array location, rather than the array index is zeroed. the entries in the binary 
fuse filter are shifted right one, and a zero is placed at index 0. This way, the phatom entry is always hot in the cache (hit in 50\% of queries). This yields
a scheme with 3.5 average cache misses and no branch mispredictions. 

% table
% scheme | # hash | mispredict | # cache miss
% cond   | 3.5    | 0.5        | 3.5
% zero   | 4      | 0          | 4
% phant  | 4      | 0          | 3.5

% Move to appendix
\begin{figure}[H]
    \centering
    \begin{minipage}[t]{0.32\textwidth}
        \centering\textbf{Conditional}
        \tiny
        \begin{algorithmic}
            \Function{Contains}{k}
                \State $f \gets$ \Call{Fingerprint}{k}
                \State $e_1 \gets A[$\Call{Hash1}{k}]
                \State $e_2 \gets A[$\Call{Hash2}{k}]
                \State $e_3 \gets A[$\Call{Hash3}{k}]
                \State $e_4 \gets 0$
                \If{\Call{IsFourEdge}{k}}
                    \State $e_4 \gets A[$\Call{Hash4}{k}]
                \EndIf
                \State $f' \gets e_1 \oplus e_2 \oplus e_3 \oplus e_4$
                \State \Return $f == f'$ 
            \EndFunction
        \end{algorithmic}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.32\textwidth}
        \centering\textbf{Conditional}
        \tiny
        \begin{algorithmic}
            \Function{Contains}{k}
                \State $f \gets$ \Call{Fingerprint}{k}
                \State $e_1 \gets A[$\Call{Hash1}{k}]
                \State $e_2 \gets A[$\Call{Hash2}{k}]
                \State $e_3 \gets A[$\Call{Hash3}{k}]
                \State $e_4 \gets A[$\Call{Hash3}{k}] * \Call{IsFourEdge}{k}
                \State $f' \gets e_1 \oplus e_2 \oplus e_3 \oplus e_4$
                \State \Return $f == f'$ 
            \EndFunction
        \end{algorithmic}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.32\textwidth}
        \centering\textbf{Conditional}
        \tiny
        \begin{algorithmic}
            \Function{Contains}{k}
                \State $f \gets$ \Call{Fingerprint}{k}
                \State $f \gets f \oplus A[$\Call{Hash1}{k}]
                \State $f \gets f \oplus A[$\Call{Hash2}{k}]
                \State $f \gets f \oplus A[$\Call{Hash3}{k}]
                \If{\Call{IsFourEdge}{k}}
                    \State $f \gets f \oplus A[$\Call{Hash4}{k}]
                \EndIf
                \State \Return $f == 0$ 
            \EndFunction
        \end{algorithmic}
    \end{minipage}
    \caption{Three query implementations for non-uniform binary fuse filters. The phantom scheme achieves 3.5 average cache misses with no branch mispredictions.}
    \label{fig:schemes}
\end{figure}

\input{figures/csc592-bigdata-nubff.tex}

\input{figures/csc592-bigdata-nusoe.tex}

\input{figures/csc592-bigdata-nuhg.tex}


% Inspired by random hyperparameter sweeps in deep learning, we computed em-
% pirical peelablity thresholds for 10403 non-uniform fuse graph families (see Table
% 1) with mean edge size between three and four. As probability of being pee-
% lable is a monotonic function of the edge density, we binary searched for the
% empirical thresholds. At each candidate threshold, three hypergraphs (1000000
% edges, 8192 window size) were sampled from the family. If the majority of the
% hypergraphs were peelable, the family was considered peelable at the threshold
% value. Otherwise, the family was considered not peelable at the threshold value.

% While k-uniform fuse graphs have integer mean edge sizes, they can be extended
% to any mean edge size through sharding [8]. For example, edges can be evenly
% split between a 3-uniform fuse subgraph and a 4-uniform fuse subgraph to form
% a hypergraph with mean edge size of 3.5. We found that many non-uniform fuse
% graph families have higher thresholds than sharded k-uniform fuse graphs.



% we considered non-uniform
% empirical peelability thresholds for 10403 non-uniform fuse graph families
% mean edge size between 3 and 4

% can interpolate via sharding to attain any threshold

% non-uniform hypergraph families
% pareto frontier
% optimized implementation

% limitations:
% explore additional parameteratizations of hg families
% need to prove conjecture
% implementation did not perform well

TODO: A 400–500-word summary of the technical contributions of the work, what was
accomplished overall, and the limitations of their work.

\subsection{Reflection}

TODO: A 250-word self reflection of what skills were developed or enhanced in completing those
projects.