\section{Entropy-Aware Hashing}

This project was completed for CSC 519: Computer Networks.

\subsection{Background}

% relevance of hash functions
% limitations of current hash functions (general)
Hashing is ubiquitous in computing. For example, routing table lookups, intrusion detection systems, load distribution, software defined networking, network 
address translation and application-layer caches rely on fast hashing~\cite{kurose2025}. Because hashing is already highly optimized for per-byte throughput, 
performance gains increasingly come from specialization, where hash functions are tailored to specific workloads. \\


% entropy learned hashing, sede
% main idea(s): hash fewer bytes, use efficient instructions
% research questions:
% - ELH + sede do not use some information: dist of keys, data structure
% incorporate this
\noindent
Recent works pursue this by exploiting the entropy structure of input data. Entropy learned hashing~\cite{hentschel2022} and \texttt{Sede}~\cite{hoffmann2025} 
automatically identify and hash only the high-entropy bytes of keys, reducing unnecessary computation. This approach is theoretically grounded: Mitzenmacher and 
Vadhan~\cite{mitzenmacher2008} showed that standard hash functions operate by extracting entropy from inputs, so hashing low-entropy bytes wastes work without 
improving hash quality. \\

\noindent
However, both methods specialize incompletely. They treat all keys as equally likely and optimize purely for collision count, ignoring non-uniform key access 
frequencies and the probe behavior of the target data structure. This potentially misses performance gains for workloads with skewed key distributions and
data structure dependent query costs. Therefore, we ask:
\begin{enumerate}
    \item How can hash functions be automatically specialized to the input distribution, target data structure and underlying hardware architecture 
    simultaneously?
\end{enumerate}

\subsection{Summary}

\subsubsection{Entropy-Aware Hashing}

We propose entropy-aware hashing, a framework that automatically specializes a hash function for a given dataset, data structure, and architecture. The 
framework has two configurable components: the hash function form which determines how data can be read and mixed, and the cost model, which defines what it 
means for a hash function to be good for a particular workload. \\

\noindent
To select which bytes of a key to hash, we extend entropy learned hashing's byte level selection to also support word and double-word granularity. Coarser 
granularity may include some lower-entropy bytes, but enables more efficient instruction selection, including SIMD and AES-NI instructions that operate on wider
data. We also expose alignment as a configurable option, since some architecture penalize unaligned memory accesses. The mixing function is drawn from: 
MurmurHash3, Thomas Wang's hash, Stafford's hash, XXHash, GXHash, and a single AES round. \\

\noindent
We evaluate four cost models. The unweighted collision cost (used by prior work) counts expected collisions uniformly across all keys. The weighted collision 
cost accounts for non-uniform access frequencies: collisions between frequently accessed keys are penalized more heavily. The maximum weighted probe cost and 
minimum weighted probe cost measure the expected probe count under the worst- and best-case chain orderings respectively, modeling the cost of collision 
resolution in separate chaining hash tables. \\ 

\noindent
To specialize, we iterate over hash function forms and greedily select data chunks to hash under the chosen cost model, then benchmark all candidates and select
the fastest. We instantiate this framework by specializing for \texttt{std::unordered\_set} (a separate chaining hash table) on a URL key dataset on the 
\texttt{x64} architecture. \\

\noindent
The fastest configuration uses unaligned double-word selection, unweighted collision cost, and a single AES round as the mixing function. This reduces hash 
latency from 12.1ns (\texttt{std::hash}) to 4.5ns, and \texttt{std::unordered\_set} query latency from 32.9ns to 14.7ns, roughly a 2$\times$ end-to-end speedup, 
with no impact on build time. Notably, specialization to key frequency distribution and data structure probe behavior did not provide a speedup: 
architecture-level specialization via unaligned double-word reads and AES-NI instructions did. 

\subsubsection{Future Work}

We see several promising directions for future work. The most fundamental is theoretical: the hardness of optimal byte subset selection for entropy 
maximization is uncharacterized, and the greedy heuristic lacks formal approximation guarantees. Establishing these would clarify whether better optimization 
algorithms are worth pursing. \\

\noindent
On the systems side, the framework currently assumes fixed-length keys and a static distribution. Extending to variable length keys is nontrivial. One promising
formulation treats alignment of high-entropy regions across keys as a multiple sequence alignment problem. Adapting to shifting distributions would require 
online entropy estimation to track high-entropy byte positions as the workload evolves. \\

\noindent
Finally, the cost model is currently hand-designed, and requires benchmarking. A learned cost model, trained from measurements across datasets, data structures,
and architectures, could automatically capture interactions that are difficult to model analytically, enabling better specialization. 

% prove hardness, greedy is approximation
% dynamicly learn entropy
% different length keys (sequence alignment)
% automatically learn cost model from DS, data and arch.

% TODO: A 400–500-word summary of the technical contributions of the work, what was
% accomplished overall, and the limitations of their work.

\subsection{Reflection}

This project developed two distinct technical competencies. The first is low-level performance optimization. To identify efficient mixing primitives, I read 
Intel's \texttt{x64} instruction and optimization reference manuals. This led me to the AES-NI extension which includes a single cycle AES round instruction. 
Learning to reason directly from architecture documentation was a meaningful shift in my optimization approach. \\

\noindent
The second is the theory of approximation algorithms. I attempted to prove that finding the maximum-entropy byte subset is \textbf{NP-Hard} and bound its 
best approximation factor via an approximation-preserving reduction. I also tried to show that the greedy algorithm achieves an optimal approximation ratio. I 
did not succeed at either. Regardless, the experience was valuable: I worked through standard reduction techniques carefully enough to understand where I 
failed, and learned about approximation while trying. I hope to continue to build my theoretical toolkit, and eventually resolve this problem.


% TODO: A 250-word self reflection of what skills were developed or enhanced in completing those
% projects.