\section{Tuning Secure Learned Bloom Filters}

This project was completed for CSC 592: Machine Learning Security.

\subsection{Background}

% AMQ filters, motivation for them + motivation for security
Database systems often require fast and small set data structures. For example, RocksDB~\cite{dong2021} reduces IO operations by filtering queries before 
reading from disk. As false positives do not impact correctness, they are permitted with bounded probability. Such approximate membership queries can be much
faster and more memory efficient than exact membership queries.  \\

\noindent
A Bloom filter~\cite{bloom1970} is a probabilistic data structure for approximate membership queries. Membership information is stored in an array of bits. To
query the filter, a key is hashed to several entries in the array (see Figure~\ref{fig:csc592_mlsec_bloom}). If all entries are set to $1$, then the key 
\emph{might} be in the set, and false positives occur with probability $\epsilon$. Otherwise, the query \emph{definitely} is not in the set. \\

\input{figures/csc592-mlsec-bloom.tex}

\noindent
Learned Bloom filters~\cite{kraska2018} extend Bloom filters with a machine learning model that predicts set membership (see Figure~\ref{fig:csc592_mlsec_lbf}). 
For many practical datasets, the model compresses membership information more efficiently than a Bloom filter, reducing memory consumption. Queries are first 
checked with the model. On a positive prediction, the key may or may not be in the set, so false positives occur at the model's false positive rate. On a 
negative prediction, the key may still be in the set due to model false negatives, so a backup Bloom filter storing all false negatives is queried. False 
positives then occur at the product of the model's true negative rate and the backup filter's false positive rate. The overall false positive rate is the sum of
these two terms. \\

% learned bloom filter diagram
\input{figures/csc592-mlsec-lbf.tex}

% downtown bodgeda filter
\noindent
Since standard machine learning models are vulnerable to adversarial attack, an attacker can generate misclassified non-members and arbitrarily degrade the 
false positive rate. Secure learned Bloom filters~\cite{almashaqbeh2026} defend against this by adding a second backup Bloom filter to confirm positive 
predictions against all true set members, rejecting model false positives. However, the backup filters are vulnerable to attack. Secure Bloom 
filters\footnote{Secure Bloom filters are often called Naor-Eylon filters.}~\cite{almashaqbeh2026} address this by applying a pseudorandom permutation to keys 
before querying, ensuring no efficient adversary can non-negligibly degrade the false positive rate (assuming $\mathbf{P} \neq \mathbf{NP}$). Secure learned 
Bloom filters therefore replace both backup filters with secure Bloom filters. The false positive rate is then bounded by the maximum of the two backup filters'
false positive rates. \\

% downtown bodgeda filter diagram
\input{figures/csc592-mlsec-dbf.tex}

\noindent
Like learned Bloom filters, several secure learned Bloom filter parameters must be tuned to attain desired expected and worst-case false positive rates, 
including the decision threshold and backup filter false positive rates. These parameters also control memory consumption, so they must be set carefully to 
preserve memory efficiency. Tuning procedures for learned Bloom filters do not directly translate to secure learned Bloom filters, because they only optimize 
for expected false positive rate, ignoring the worst-case guarantee. Therefore, we ask:
\begin{enumerate}
    \item How can the decision threshold and backup filter false positive rates be chosen to minimize memory footprint given desired expected and maximum false 
    positive rates?
    \item How well do secure learned Bloom filters perform in practice with and without parameter tuning?
\end{enumerate}

\subsection{Summary}

% research questions
% optimal backup filter sizing algorithm
% empirical attack on DBF

\subsubsection{Tuning Algorithm}

Given a desired expected false positive rate $\epsilon$ and maximum false positive rate $\epsilon_\textrm{max}$, we wish to set the decision threshold $\tau$ 
and backup filter false positive rates such that the filter size is minimized. We proceed in two steps. First, fix the decision threshold, and find backup 
filter false positive rates that yield the smallest filter obeying the expected false positive rate. Second, choose the decision threshold that yields the 
smallest filter obeying the maximum false positive rate. \\

\noindent
The total size of a secure learned Bloom filter in bits is 
\begin{align*}
    - \frac{n}{\ln(2)^2}\left( \text{M}_\text{TPR} \ln(\text{TP}_\text{FPR}) + \text{M}_\text{FNR} \ln(\text{FN}_\text{FPR})  \right) + \text{M}_\text{bits}
\end{align*}
where $n$ is the set's size, $\mathrm{M}$ is the model, $\mathrm{TP}$ is the true positive filter and $\mathrm{FN}$ is the false negative filter. 
$\text{M}_\text{bits}$ is the memory consumption of the learned model in bits. Subscripts on $\text{M}$ denote model rates, and subscripts on $\text{TP}$ and 
$\text{FN}$ denote backup filter false positive rates. \\

\noindent
Given a fixed decision threshold, we prove the total size is minimized subject to expected false positive rate $\epsilon$ by taking 
\begin{align*}
   \mathrm{TP}_\mathrm{FPR} = \frac{\mathrm{M}_\mathrm{TPR}}{\mathrm{M}_\mathrm{FPR}} \cdot \epsilon \qquad \mathrm{FN}_\mathrm{FPR} = \frac{\mathrm{M}_\mathrm{FNR}}{\mathrm{M}_\mathrm{TNR}} \cdot \epsilon
\end{align*}
where $\mathrm{M}$ is the model, $\mathrm{TP}$ is the true positive filter, and $\mathrm{FN}$ is the false negative filter. Subscripts denote false positive,
false negative, true positive, and true negative rates. \\

\noindent 
Then, we show that the optimal choice of decision threshold minimizes 
\begin{align*}
    C(\tau) = -\text{M}_\text{TPR} \ln \left( \frac{\text{M}_\text{TPR}}{\text{M}_\text{FPR}} \right) 
 - \text{M}_\text{FNR} \ln \left( \frac{\text{M}_\text{FNR}}{\text{M}_\text{TNR}}\right)
\end{align*}
To maintain worst-case security guarantees (i.e. bounded maximum false positive rate), we restrict the choice of $\tau$ such that 
\begin{align*}
    \max\left\{\frac{\text{M}_\text{TPR}}{\text{M}_\text{FPR}} \cdot \epsilon, \frac{\text{M}_\text{FNR}}{\text{M}_\text{TNR}} \cdot \epsilon  \right\} \leq \epsilon_\text{max}
\end{align*}
so the false positive rate remains bounded even if the adversary forces all-positive or all-negative predictions. The optimal $\tau$ can be found in 
$\Theta(n \lg n)$ time (see Algorithm~\ref{fig:csc592_mlsec_alg}). \\

\input{figures/csc592-mlsec-alg.tex}

\subsubsection{Tuned Performance}

We construct a secure Bloom filter and a secure learned Bloom filter from a dataset of malicious and benign 
URLs\footnote{\url{https://www.kaggle.com/datasets/sid321axn/malicious-urls-dataset}}. For the same memory budget and maximum false positive rate, the tuned 
secure learned Bloom filter reduces the expected false positive rate by half relative to the secure Bloom filter (9.7\% vs. 4.6\%). We verify our implementation
using several adaptive adversarial attacks, and find none increase the false positive rate above $\epsilon_\mathrm{max}$. 

\subsubsection{Future Work}

We see three promising directions for future work. First, we should compare the tuned secure learned Bloom filter against an untuned secure learned Bloom 
filter, not just a secure Bloom filter. Second, no efficient implementation of the secure learned Bloom filter exists, so its real-world relevance remains 
unknown. Third, our tuning algorithm should adapt to different constraint configurations: for example, minimizing expected false positive rate subject to both 
a memory budget and a maximum false positive rate.  

\subsection{Reflection}

% TODO: A 250-word self reflection of what skills were developed or enhanced in completing those
% projects.

This project strengthened my ability to enter a new subfield independently. I attended several research talks on machine learning security and privacy at Brown 
University. Although I had no cryptography background, I learned to identify areas where my existing expertise could contribute and where I would be out of my 
depth. I evaluated several potential projects, recognized that most required background I lacked, and narrowed my focus to tuning secure learned Bloom filters. 
My experience in machine learning, approximate membership filters and algorithm design made this topic approachable. \\

\noindent
This project also developed my appreciation for upfront scoping. Working under a two-week timeline, I did not define objectives or evaluation criteria before 
starting. As a result, I missed that my tuning algorithm did not address the most important constraint configuration, and that my baseline provided only partial
evidence for my claims. Both oversights are now documented as future work, but a clear plan would have helped me catch these missteps sooner. \\

\noindent
Finally, this project enhanced my ability formalize and solve optimization problems. I identified and formalized the relevant constraints and objectives, then
decomposed the problem into two subproblems. I solved one with calculus and the other with dynamic programming. On its own, each step was straightforward, but
identifying the correct steps to take was a challenge.