\section{Non-Uniform Binary Fuse Filters}

This project was completed for CSC 592: Algorithms for Big Data.

\subsection{Motivation}

Distributed systems often require fast and small set data structures. For example, content delivery networks avoid caching one-hit wonders by querying a visited
set~\cite{maggs2015}. Supporting such exact membership queries on a large universe\footnote{Set of all keys.} $U$ requires $\Omega(\lg |U|)$ bits per key. For 
large enough sets, this is impractical. By permitting false positive results with some probability $\epsilon$, the lower bound is reduced to 
$\lg \frac{1}{\epsilon}$ bits per key~\cite{carter1978}. In other words, \textit{approximate} membership queries can be much more memory efficient than exact 
membership queries~\cite{bloom1970, graf2020, graf2022}. \\

\noindent
A binary fuse filter~\cite{graf2022} is a recent static probabilistic data structure for approximate membership queries. Membership information is stored in an 
array of $m$-bit values partitioned into disjoint segments. To query the filter, a key is hashed to a fingerprint value and to three array entries located in 
consecutive segments (see Figure~\ref{fig:csc592_bigdata_bff}). If the bitwise-XOR of the entries is equal to the fingerprint, then the query \emph{might} be in 
the set. Otherwise, the query \emph{definitely} is not in the set.

\input{figures/csc592-bigdata-bff.tex}

\noindent
To construct a binary fuse filter, choose each array entry such that, for each key, the bitwise-XOR of the three corresponding array entries is equal to
the fingerprint. This requires solving a system of linear equations with variables in $\mathbb{F}^{m}_2$ (i.e. $m$-bit vectors) where addition is performed 
using bitwise-XOR. Let $h_1$, $h_2$ and $h_3$ be the hash functions, $f$ be the fingerprint function, and $A$ be the array. Each key $k$ yields a linear 
equation
\begin{align*}
    A[h_1(k)] \oplus A[h_2(k)] \oplus A[h_3(k)] = f(k)
\end{align*}
that must be satisfied (see Figure~\ref{fig:csc592_bigdata_soe}). If some equation is not satisfied, the filter will incorrectly report the corresponding key is
\emph{definitely} not in the set. Since approximate membership queries allow only false positives and not false negatives, the choice of array entries must 
satisfy every equation.

\input{figures/csc592-bigdata-soe.tex}

\noindent
To solve the system, model it as a hypergraph. A hypergraph is a graph where edges may contain any subset of vertices. For each entry, create a vertex. For each 
equation, create an edge containing entries (vertices) in the equation (see Figure~\ref{fig:csc592_bigdata_hg}). Then, 2-peel the hypergraph, that is:
\begin{enumerate}
    \item Remove vertices of degree zero, then set their entries to $0^m$.
    \item Remove vertices of degree one and their incident edges. Then, set their entries to the bitwise-XOR of the fingerprint and the two other entries in the
    edge. If the other entries are not yet set, defer until they are set.
    \item Repeat until there are no vertices of degree less than two, or until the empty hypergraph is reached (i.e. all entries set).
\end{enumerate}
If the empty hypergraph is reached, the hypergraph is called 2-peelable. The system is solved if and only if its hypergraph is 2-peelable. \\

\input{figures/csc592-bigdata-hg.tex}

\noindent
In the filter setting, edges represent keys and vertices represent array entries. Lower edge size means cheaper queries, since fewer array entries (vertices) 
are accessed per key (edge). Higher edge density means lower space usage, since there are more keys (edges) per array entry (vertex). \\

\noindent
Many hypergraph families exhibit sharp phase transitions in peelability~\cite{dietzfelbinger2019, walzer2021}. When the edge density is below a critical value, 
called the 2-peelability threshold, sampled hypergraphs are almost surely 2-peelable. When the density exceeds this threshold, they are almost surely not. The 
threshold is therefore a direct measure of space efficiency. A higher threshold means more keys can be stored per array entry. \\

\noindent
Binary fuse filters induce the fuse graph family, which achieves the supremum 2-peelability threshold among 3-uniform hypergraph\footnote{In general, a 
hypergraph is $m$-uniform if every edge contains exactly $m$ vertices.} families~\cite{walzer2021}. This threshold occurs at an edge density of $0.918$, or 
roughly $1.089 \lg \frac{1}{\epsilon}$ bits per key~\cite{walzer2021}. No family of 3-uniform hypergraphs has higher 2-peelability threshold. More generally, a 
binary fuse filter with $m \geq 3$ hash functions\footnote{Only binary fuse filters with three or four hash functions are practical. With four hash functions, 
the 2-peelability threshold is already $0.977$, yielding about $1.023 \lg \frac{1}{\epsilon}$ bits per key or within $2.3\%$ of optimal~\cite{walzer2021}.} 
saturates the 2-peelability threshold for $m$-uniform hypergraphs~\cite{walzer2021}. \\

\noindent
However, this optimality result is confined to uniform hypergraphs. There may exist families of non-uniform hypergraphs with higher 2-peelability thresholds. If
such families also have a low average edge size (i.e. few hashes and accesses per query), filters based on them could potentially reduce memory overhead without
degrading query latency and throughput. Therefore, we ask:

\begin{enumerate}
    \item How can we systematically construct non-uniform hypergraph families with peelability thresholds comparable to or surpassing those of fuse graph 
    families?
    \item How can we design efficient approximate membership query data structures by exploiting peelable non-uniform hypergraph families?
\end{enumerate}

\subsection{Summary}

\subsubsection{Non-Uniform Hypergraphs}
\label{sec:nuhg}

We computed 2-peelability thresholds for 10K non-uniform fuse graph families with mean edge size between 3 and 4 (see 
Figure~\ref{fig:csc592_bigdata_nuhg_graph}). As peelability is monotonic in edge density, we used binary search. At each candidate threshold, we sampled 
three hypergraphs with 1M edges. If the majority were peelable, we considered the family peelable at that threshold. We implemented this algorithm in C++ and 
found thresholds in parallel on a compute cluster. \\

\input{figures/csc592-bigdata-nuhg-graph.tex}

\noindent
While $m$-uniform graphs have integer mean edge sizes, they can be extended to any mean edge size via sharding\footnote{Sharding partitions a dataset into 
disjoint subsets. For hypergraphs, each subset becomes an edge-disjoint subgraph. For filters, each subset is stored in a separate filter.}~\cite{graf2022}. For
example, splitting edges evenly between edge-disjoint 3-uniform and 4-uniform subgraphs yields a mean edge size of 3.5. Many non-uniform families outperform 
this baseline, and therefore afford smaller filters. Interestingly, some non-uniform families also outperform uniform families at integer mean edge sizes. Since 
peelability thresholds are defined in the limit of large graphs, a non-uniform family can be more peelable in practice than a uniform family with nearly the 
same mean edge size. It's also possible three samples per candidate is not enough to accurately determine the threshold. For example, one family achieves 
$\approx 1$\% memory savings over the uniform baseline at mean edge size $4$. \\

\noindent
These data suggest two conjectures. Let $\mathcal{H}_d$ denote the set of fuse graph families with mean edge size $d$ and let $\tau(\mathcal{F})$ 
denote the 2-peelability threshold of a family $\mathcal{F}$. Define $\Gamma(d) = \sup \{ \tau(\mathcal{F}) : \mathcal{F} \in \mathcal{H}_d \}$. We conjecture 
that $\Gamma(d)$ is continuous, monotonically increasing, and concave. Since a concave $\Gamma$ would lie above any linear interpolation, this implies that 
sharding is suboptimal. We further conjecture that for $d \not \in \mathbb{Z}$, the supremum is attained by the family with a fraction $(1 - p)$ of 
$\lfloor d \rfloor$-edges and a fraction $p$ of $\lceil d \rceil$-edges, where $p = d - \lfloor d \rfloor$. For example, for $d = 3.5$, this family contains
50\% 3-edges and 50\% 4-edges. Both conjectures remain open.

\subsubsection{Non-Uniform Binary Fuse Filters}

\noindent
We implemented and optimized a non-uniform binary fuse filter with 50\% 3-edges and 50\% 4-edges in C++ (i.e. each key hashes to either 3 or 4 array entries). 
This edge distribution offered the greatest advantage over sharding uniform binary fuse filters. We considered three query algorithms: conditional branch, 
branchless and phantom entry described below (see Table~\ref{tab:csc592_bigdata_impl_tab} for metrics and Section~\ref{sec:csc592_bigdata_contains_impl} for 
pseudocode). \\

\input{figures/csc592-bigdata-impl-tab}

\noindent
In all three algorithms, we report that the key \emph{might} be in the set if the XOR of the accessed array entries equals the fingerprint. The algorithms 
differ only in how they handle variable edge size when computing the XOR. \\

\noindent
In the conditional branch algorithm (see Figure~\ref{fig:csc592_bigdata_pc_cond}), we XOR the first three array entries. If the key is a 4-edge, we XOR in the 
fourth entry. This branch mispredicts on at least half of all queries.  \\

\noindent
In the branchless algorithm (see Figure~\ref{fig:csc592_bigdata_pc_bl}), we load all four array entries unconditionally. If the key is a 3-edge, we zero the 
fourth value with a mask before XORing all four values. For large filters, all four loads will miss the cache. \\ 

\noindent
In the phantom entry algorithm (see Figure~\ref{fig:csc592_bigdata_pc_phant}), we load all four array entries unconditionally. However, if the key is a 3-edge, 
we zero the fourth hash address before the load, redirecting it to $A[0]$. To support this, we shift all valid entries left by one index and set $A[0] = 0$ at
construction time (see Figure~\ref{fig:csc592_bigdata_nubff}). Unlike the branchless algorithm, the fourth load for 3-edges hits the cache, since $A[0]$ is 
accessed in 50\% of queries. \\

\input{figures/csc592-bigdata-nubff.tex}

\noindent
Surprisingly, only the conditional branch algorithm achieved higher query throughput than the 4-uniform binary fuse filter, and even this fell short of an 
acceptable time-space tradeoff. Since we expected nearly all cycles to be spent stalled on memory loads, reducing cache misses per query should have improved 
throughput proportionally. In practice, however, only 80\% of cycles were stalled on loads, and there were no other stalls. \\

\noindent
We hypothesize that out-of-order (OoO) execution stalls before it can compute load addresses for the next key. As a result, when outstanding loads complete, 
hashes for the next query have not yet been computed, and the subsequent loads are delayed by the full hash latency. This dependency also explains the 
conditional branch algorithm's advantage: its lower hash latency reduces this delay. \\

\noindent
The root cause is likely one of two structural hazards: either a shortage of miss status holding registers (MSHRs) prevents prefetching the next key (and 
therefore precomputing its hashes), or the reorder buffer (ROB) is too small to precompute enough hashes for future keys. Based on microarchitectural 
parameters and the benchmark assembly, the first explanation is more likely.

\subsubsection{Future Work}

We see four promising directions for future work. First, batching may yield a non-uniform binary fuse filter with a competitive time-space tradeoff. Second, 
interleaving the segments of a binary fuse filter could improve cache locality during queries. Third, introducing additional weak points into the fuse graph 
would enable greater parallelism during filter construction. Currently, the fuse graph has only two weak points, one at each end, limiting peeling to two 
threads. The recommended workaround is sharding, but this may sacrifice memory efficiency. Fourth, we aim to prove the conjectures stated in 
Section~\ref{sec:nuhg}, for which the methods of \cite{dietzfelbinger2019} and \cite{walzer2021} may be useful.

\subsection{Reflection}

This project strengthened my ability to enter a new subfield independently through a literature review. Reading many approximate membership query papers, from 
theoretical analysis to engineered implementations, gave me enough grounding to pose original research questions. One area I want to develop further is posing
robust questions, where even a negative result is informative. If non-uniform filter's microarchitectural bottleneck cannot be overcome, that facet is a total 
loss. \\

\noindent
The most demanding aspect of this project was maintaining multiple representations of the same structure simultaneously: a filter is a data structure, a system
of linear equations, and a hypergraph. Initially, I found it challenging to move fluidly between these views. Over time this became natural, and I applied it 
most directly when using cluster resources to empirically explore non-uniform hypergraphs, and translate the result into both formal conjectures and a concrete
implementation. \\

\noindent
The performance engineering work expanded my understanding of modern processor architecture. Attempting to make the non-uniform filter practical required 
reasoning carefully about superscalar out-of-order execution and non-blocking caches, interpret hardware performance counters, and diagnose microarchitectural 
bottlenecks. \\

\noindent
Finally, I want to thank Morgan, Tom, Vincent and Dan. Collaborating with them is definitely a highlight of my academic experience.