\section{Introduction}

The projects in this portfolio grew out of self-directed study. Unlike my published research, which developed within an established advising relationship and 
with relevant coursework as a foundation, this work required me to acquire the necessary background, conduct literature reviews, and define and pursue 
research questions independently. Each project addresses a distinct problem in performance engineering of hash-based data structures and hash functions. For 
each, I generalized a restrictive parameterization, then automatically optimized within the broader space to improve performance. \\

\noindent
The first generalizes binary fuse filters, a recent approximate set membership structure, to support non-integer arity\footnote{The number of hash functions per
query.} natively. Binary fuse filters offer only a discrete tradeoff: increasing arity reduces memory overhead but increases query time. The accepted workaround
partitions data between disjoint filters of different arity to achieve a continuous tradeoff. Native non-integer arity makes this principled, and 
implementations found by large-scale search require less memory than the partitioning approach. \\

\noindent
The second introduces the first optimal parameter tuning algorithm for secure learned Bloom filters, a recent approximate set membership structure that combines
a learned model with two secure filters. Since no tuning algorithms were known, the recommended approach was simply to allocate memory evenly between the
two filters. This default cannot target an expected false positive rate, nor bound the worst-case rate, and wastes memory when either constraint is defined. I 
give the first exact tuning algorithm to minimize memory footprint subject to expected and maximum false positive rate constraints. \\

\noindent
The third extends entropy-aware hash function specialization, a technique to select just enough bytes to form a good enough hash. Previous cost models ignored 
key frequency, target data structure and target architecture, leading to incomplete specialization and suboptimal performance. I generalize cost models to 
incorporate these factors, then demonstrate the approach by specializing for \texttt{std::unordered\_set} with URL keys on the \texttt{x64} platform. The 
resulting hash function reduces both hash and hash table query latency. \\

\noindent
Together these projects reflect a single approach to performance engineering: define expressive parameter spaces and automatically optimize within them. Each 
project required independently building fluency in an unfamiliar area, and each left open problems I continue to pursue. They are the clearest evidence that I 
have become a computer scientist.